Conversational expectations account for apparent limits on theory of mind use

### 2017-11-13 writing/journal_manuscript/pnas-submission.tex
Minor edits to introduction

% Characterize the computational approach to theory of mind we plan to take
% Point out a gaping hole that could benefit from this approach
Our success as a cooperative species depends upon our ability to make sense of other agents. 
A natural solution to this problem is to maintain an internal predictive model of how minds work: how an agent's unobservable beliefs and desires give rise to observable actions. 
Armed with a computational theory of mind, people can make inferences about an agent's likely mental state given their past behavior and run simulations to form sophisticated expectations about their future behavior. 
This theoretical approach to social cognition, where theory of mind reasoning \cite{PremackWoodruff78_ChimpanzeeToM, GopnikWellman92_TheoryTheory, WellmanCrossWatson01_ToMMetaAnalysis, SetohScottBaillargeon16_FalseBelief, ScottBaillargeon17_EarlyToM} is grounded in an explicit computational model of how other minds work, has been useful for understanding a number of experimental phenomena in both adults and children \cite{KileyHamlinEtAl13_MentalisticCoreSocial, PantelisEtAl14_IntentionInference, BakerSaxeTenenbaum09_ActionUnderstandingInversePlanning, JaraEttingerEtAl16_NaiveUtilityCalculus, BakerEtAl17_ToMBayesian}. 
Yet our current understanding of theory of mind use in \emph{communication} remains limited. 
Theory of mind as applied within communication depends on an accurate model of what a speaker would say, or what a listener would think, in different situations. 
Past efforts have instead largely focused on a relatively narrow behavioral operationalization of theory of mind as perspective-taking about a partner's visual access \cite{KeysarBarr___Brauner00_TakingPerspective, KeysarLinBarr03_LimitsOnTheoryOfMindUse, LinKeysarEpley10_ReflexivelyMindblind}.

%While these failures are compelling evidence against a simple model where theory of mind reasoning is limited to literal perspective-taking, 
We argue in this paper that the mental models supporting theory of mind use in communication are much richer than past efforts have explored. 
In particular, we consider a class of mental models representing not only a partner's visual access but also Gricean expectations of parsimony and cooperativity \cite{Grice75_LogicConversation}: that speakers will avoid saying things that are confusing or unnecessarily complicated given the current context. 
For example, suppose a friend wants to refer to a nearby Dalmatian. 
%There are numerous descriptions that would all be literally true, but different expressions are preferred in different contexts.
If there are no other dogs nearby, we may expect them to use a parsimonious basic-level term like ``dog.''
When there is potential for confusion, however, we expect them to consider our likely interpretation and introduce additional modifiers (``spotted dog'') or shift to more specific terms (``Dalmatian'') to disambiguate the one they are trying to refer to \cite{BrennanClark96_ConceptualPactsConversation, VanDeemter16_ComputationalModelsOfReferring, GrafEtAl16_BasicLevel}. 
When speakers violate these expectations, listeners draw inferences about the intended meaning in order to account for the violation: for example, XXX \cite{XXX, BrehenyKatsosWilliams06_ImplicaturesOnline, StillerGoodmanFrank15_AdHocImplicature}.

@@

The most immediate model of social reasoning is consideration of the informativeness of a label in a shared context containing many objects. For a speaker, this means preferring labels that are good fits for the target but \emph{not} known distractors in common view. For instance, when referring to a dog in the shared context of other dogs, a speaker may opt to be more informative than simply saying `dog,' since it is ambiguous. For a listener, this means discounting objects that are not in shared context, regardless of how well the description fits. These are not available to the visual access of the speaker and thus not viable referents.

>>>>

%%% Characterize the computational approach to theory of mind we plan to take
%%% Point out a gaping hole that could benefit from this approach
Our success as a cooperative species depends upon our ability to make sense of other agents. 
One solution is to maintain an internal predictive model of how minds work: how an agent's unobservable beliefs and desires give rise to observable actions. 
Armed with a computational theory of mind, we can make inferences about an agent's likely mental state given their past behavior and form sophisticated expectations about their future behavior. 
%This theoretical approach to social cognition, where theory of mind reasoning \cite{PremackWoodruff78_ChimpanzeeToM, GopnikWellman92_TheoryTheory, WellmanCrossWatson01_ToMMetaAnalysis, SetohScottBaillargeon16_FalseBelief, ScottBaillargeon17_EarlyToM} is grounded in an explicit computational model of how other minds work, has 
% The approach has been useful for understanding a number of experimental phenomena in both adults and children \cite{KileyHamlinEtAl13_MentalisticCoreSocial, PantelisEtAl14_IntentionInference, BakerSaxeTenenbaum09_ActionUnderstandingInversePlanning, JaraEttingerEtAl16_NaiveUtilityCalculus, BakerEtAl17_ToMBayesian}. 
Yet our current understanding of theory of mind use in \emph{communication} remains limited. 
Just as making sense of an agent's physical movements in the world requires an accurate model of how beliefs and intentions translate into motor plans, making sense of language depends on an accurate model of what a speaker would say, or what a listener would think, in different situations.

%While these failures are compelling evidence against a simple model where theory of mind reasoning is limited to literal perspective-taking, 
Past efforts have focused on a relatively narrow behavioral operationalization of theory of mind as perspective-taking about a partner's visual access \cite{KeysarBarr___Brauner00_TakingPerspective, KeysarLinBarr03_LimitsOnTheoryOfMindUse, LinKeysarEpley10_ReflexivelyMindblind}. 
We argue in this paper that the mental models supporting theory of mind use in communication are much richer than previously explored. 
In particular, we consider a class of mental models representing not only a partner's visual access but also Gricean expectations of parsimony and cooperativity \cite{Grice75_LogicConversation}: that speakers will avoid saying things that are confusing or unnecessarily complicated given the current context. 
For example, suppose a friend wants to refer to a nearby Dalmatian. 
%There are numerous descriptions that would all be literally true, but different expressions are preferred in different contexts.
If there are no other dogs nearby, we may expect them to use a parsimonious basic-level term like ``dog.''
When there is potential for confusion, however, we expect them to consider our likely interpretation and introduce additional modifiers (``spotted dog'') or shift to more specific terms (``Dalmatian'') to disambiguate the one they are trying to refer to \cite{BrennanClark96_ConceptualPactsConversation, VanDeemter16_ComputationalModelsOfReferring, GrafEtAl16_BasicLevel}. 
When speakers violate these expectations, listeners draw inferences about the intended meaning in order to account for the violation: for example, XXX \cite{XXX, BrehenyKatsosWilliams06_ImplicaturesOnline, StillerGoodmanFrank15_AdHocImplicature}.

@@

The most immediate model of social reasoning is consideration of the informativeness of a label in a shared context containing many objects. For a speaker, this means preferring labels that are good fits for the target but \emph{not} known distractors in common view. For instance, when referring to a dog in the shared context of other dogs, a speaker may opt to be more informative than simply saying `dog,' since it is ambiguous. For a listener, this means discounting objects that are not in shared context, regardless of how well the description fits. These are not available to the visual access of the speaker and thus not viable referents. Note that
